# Modeling experimental data with linear mixed-effects models

## Learning objectives

* specify linear mixed-effects models for experimental designs
* simulate data for a design with crossed random factors of subjects and stimuli
* estimate parameters for linear mixed-effects models with crossed random factors

## Web app

- [Demo of crossed random effects](http://shiny.psy.gla.ac.uk/Dale/crossed)

## Replacing t-test and ANOVA with linear mixed-effects regression

As we have been emphasizing throughout this course, most data in psychology is multi-level---there are multiple observations on the DV for each subject. 

In the last chapter, we saw how to perform multilevel linear regression with a continuous predictor. In this section, we'll see how a mixed-effects model can be used to replace a wide variety of standard techniques used in psychology for multi-level data with categorical predictors. Indeed, a very simple model with random intercepts can replace the following techniques, if they are used on **multi-level data**:

* one sample t-test
* between-subjects designs
    - independent samples t-test
    - one factor or factorial ANOVA
* within-subjects designs
    - paired samples t-tests
    - one factor repeated measures ANOVA
    - mixed-design ANOVA
    - within-subjects factorial ANOVA

### Example: Independent-samples $t$-test on multi-level data

Let's consider a situation where you are testing the effect of alcohol consumption on simple reaction time (e.g., press a button as fast as you can after a light appears). To keep it simple, let's assume that you have collected data from 14 participants randomly assigned to perform a set of 10 simple RT trials after one of two interventions: drinking a pint of alcohol (treatment condition) or a placebo drink (placebo condition).  You have 7 participants in each of the two groups. Note that you would need more than this for a real study.

The web app below presents simulated data from such a study. Subjects P01-P07 are from the placebo condition, while subjects T01-T07 are from the treatment condition.

```{r icc-app, echo=FALSE, fig.cap="Multi-level data from an independent samples design."}
knitr::include_app("http://shiny.psy.gla.ac.uk/Dale/icc", height = "620px")
```

If we were going to run a t-test on these data, we would first need to calculate subject means, because otherwise the observations are not independent. You could do this as follows. (If you want to run the code below, you can download sample data from the web app above and save it as `independent_samples.csv`).

```{r subj-means, message = FALSE}
library("tidyverse")

dat <- read_csv("data/independent_samples.csv", col_types = "cci")

subj_means <- dat %>%
  group_by(subject, cond) %>%
  summarise(mean_rt = mean(RT)) %>%
  ungroup()

subj_means
```

Then, the $t$-test can be run using the "formula" version of `t.test()`.

```{r run-t}
t.test(mean_rt ~ cond, subj_means)
```

While there is nothing wrong with this analysis, aggregating the data throws away information. we can see in the above web app that there are actually two different sources of variability: trial-by-trial variability in simple RT (represented by $\sigma$) and variability across subjects in terms of their how slow or fast they are relative to the population mean ($\gamma_{00}$).  The Data Generating Process for response time ($Y_{st}$) for subject $s$ on trial $t$ is shown below.

*Level 1:*

\begin{equation}
Y_{st} = \beta_{0s} + \beta_{1} X_{s} + e_{st}
\end{equation}

*Level 2:*

\begin{equation}
\beta_{0s} = \gamma_{00} + S_{0s}
\end{equation}

\begin{equation}
\beta_{1} = \gamma_{10}
\end{equation}

*Variance Components:*

\begin{equation}
S_{0s} \sim N\left(0, {\tau_{00}}^2\right) 
\end{equation}

\begin{equation}
e_{st} \sim N\left(0, \sigma^2\right)
\end{equation}

In the above equation, $X_s$ is a numerical predictor coding which condition the subject $s$ is in; e.g., 0 for placebo, 1 for treatment.

The multi-level equations are somewhat cumbersome for such a simple model; we could just reduce levels 1 and 2 to 

\begin{equation}
Y_{st} = \gamma_{00} + S_{0s} + \gamma_{10} X_s + e_{st},
\end{equation}

but it is worth becoming familiar with the multi-level format for when we encounter more complex designs.

Unlike the `sleepstudy` data seen in the last chapter, we only have one random effect for each subject, $S_{0s}$. There is no random slope. Each subject appears in only one of the two treatment conditions, so it would not be possible to estimate how the effect of placebo versus alcohol varies over subjects.  The mixed-effects model that we would fit to these data, with random intercepts but no random slopes, is known as a **random intercepts model**.

A random-intercepts model would adequately capture the two sources of variability mentioned above: the inter-subject variability in overall mean RT in the parameter ${\tau_{00}}^2$, and the trial-by-trial variability in the parameter $\sigma^2$. We can calculate the proportion of the total variability attributable to individual differences among subjects using the formula below.

$$ICC = \frac{{\tau_{00}}^2}{{\tau_{00}}^2 + \sigma^2}$$

This quantity, known as the **intra-class correlation coefficient**, and tells you how much clustering there is in your data. It ranges from 0 to 1, with 0 indicating that all the variability is due to residual variance, and 1 indicating that all the variability is due to individual differences among subjects.

The lmer syntax for fitting a random intercepts model to the data is `lmer(RT ~ cond + (1 | subject), dat, REML=FALSE)`. Let's create our own numerical predictor first, to make it explicit that we are using dummy coding.

```{r dat2}
dat2 <- dat %>%
  mutate(cond_d = if_else(cond == "T", 1L, 0L))

distinct(dat2, cond, cond_d)  ## double check
```

And now, estimate the model.

```{r mod-estimate, message = FALSE}
library("lme4")

mod <- lmer(RT ~ cond_d + (1 | subject), dat2, REML = FALSE)

summary(mod)
```

Play around with the sliders in the app above and check the lmer output panel until you understand how the output maps onto the model parameters.

### When is a random-intercepts model appropriate?

The random-intercepts model is appropriate for any one-sample or between-subjects data where you have multiple observations per participant. The data **must** be multi-level, because in single-level designs the subject variability is perfectly confounded with residual variability, making it impossible for the estimation algorithm to distinguish the two sources.

A random-intercepts model is **sometimes** appropriate for within-subjects or mixed-design data. For designs where there is at least one within-subjects factor, and where you are performing analyses on subject means, it is generally the case that a random-intercepts model is appropriate.

In a design with a single within-subjects factor, it is **only** appropriate when you have a single observation per subject per level of the within-subject factor. If you have more than one observation per subject per level, you need to enrich your random effects structure with random slopes, as described in the next section. If the reason you have multiple observations per subject per level is because you have each subject reacting to the same set of stimuli, then you might want to consider a mixed-effects model with crossed random effects for subjects and stimuli, as described in the next chapter. If you have sets of individual subject means on which you would normally perform a paired-samples t-test or repeated measures ANOVA, that would be data for which the random-intercepts model would be appropriate. 

The same logic goes for factorial designs in which there is more than one within-subjects factor. In factorial designs, the random-intercepts model is appropriate if you have one observation per subject per **cell** formed by each combination of the within-subjects factors. For instance, if $A$ and $B$ are two two-level within-subject factors, you need to check that you have only one observation for each subject in $A_1B_1$, $A_1B_2$, $A_2B_1$, and $A_2B_2$. If you have more than one observation, you will need to consider including a random slope in your model.

### Rules for choosing random effects

For designs having at least one within-subject factor, it is typically the case that you need to also consider including random slopes for certain factors.  We will go into this in more detail in the next chapter, where we will consider more complex designs, but here are the basic guidelines.

For categorical factor $A$, you need to consider a random slope for $A$ in the model if **both** of the following criteria are met:

1. $A$ is a within-subjects factor;
2. Each subject has multiple observations on the dependent variable for each level of $A$.

You would specify a random slope for $A$ using the following model syntax:

`DV ~ A + (A | subject)`

where `DV` is the name of your dependent variable and `subject` is the name of the variable you use to identify subjects.

What about interactions? If you have a design with three factors, $A$, $B$, and $C$, you have three main effects, three two-way interactions ($AB$, $AC$, and $BC$), and one three-way interaction $ABC$. So there are 7 possible random slopes to consider. Which of these effects would need a random slope? Here are the steps you need to follow to figure it out for each interaction term.

1. If all factors involved in the interaction are between subjects, you don't need a slope.
2. Identify the highest-order combination of within-subject factors included in the interaction. So if $A$ and $B$ are within-subject factors and $C$ is between subjects, then $AB$ is the highest order combination. If $A$ is within subjects and $B$ and $C$ are between subjects, then $A$ is the highest order combination.
3. Check whether each subject has multiple observations on the DV in each cell (or level) resulting from the combination of the within factors you identified in the previous step. If you do, then you need a random slope. If you have only one observation per subject/cell combination, then you don't.


```{r setup, include=FALSE}
library("kableExtra")
library("tidyverse")

## paste
.p <- paste0

## .fraction
.f <- function(x, y) {
  paste0("\\frac{", x, "}{", y, "}")
}

## y-bar
.yb1 <- function(x) {
  paste0("$\\bar{Y}_{", x, "}$")
}

.yb2 <- function(x) {
  paste0("\\bar{Y}_{", x, "}")
}

## subtraction term
.st <- function(x, y, bracket = NULL) {
  if (is.null(bracket)) {
    paste0(x, " - ", y)
  } else {
    paste0(bracket[1], x, " - ", y, bracket[2])
  }
}

.rb <- c("(", ")")
.dr <- c("\\displaystyle\\left(", "\\right)")
.ds <- c("\\displaystyle\\left[", "\\right]")
```

## Specifying random effects

We'll start out with some interactive tasks where you are given data to examine and asked to specify the maximal random effects structure. The data are stored in [this zip file](data/rfx_data.zip). The archive includes four R binary files (`.rds`), which you can load in using the `readRDS()` function. Extract it and set your working directory to where the resulting subdirectory `rfx_data/` is located. 

Use the code below to load them in.

```{r load-in0, echo=FALSE}
ds1 <- readRDS("data/dataset1.rds")
ds2 <- readRDS("data/dataset2.rds")
ds3 <- readRDS("data/dataset3.rds")
ds4 <- readRDS("data/dataset4.rds")
```

```{r load-in, eval=FALSE}
ds1 <- readRDS("rfx_data/dataset1.rds")
ds2 <- readRDS("rfx_data/dataset2.rds")
ds3 <- readRDS("rfx_data/dataset3.rds")
ds4 <- readRDS("rfx_data/dataset4.rds")
```

Each object has data from a hypothetical study with a factorial design. Your task is to specify the design formula for the mixed-effects model you would fit to the data. To determine the appropriate design, you will need to inspect each dataset carefully. You want to treat subjects and stimuli as random factors. If the dataset does not include one of the random factors you can omit it from the formula.

Each dataset may have some or all of the following variables:

| Variable  | Description                                 |
|-----------+---------------------------------------------|
| `subj_id` | subject identifier                          |
| `item_id` | stimulus identifier                         |
| `A`       | a categorical factor (independent variable) |
| `B`       | a categorical factor (independent variable) |
| `C`       | a categorical factor (independent variable) |
| `DV`      | the dependent variable                      |

Your task is to determine the appropriate mixed-effects model specification including the maximal random-effects structure, and write the `formula` part of the `lme4::lmer()` model as your answer.  For all of the models, the variable `DV` should appear on the left-hand side of the formula; i.e., it should be of the form `DV ~ A + ...`.  If a given dataset includes both `subj_id` and `item_id`, you need to specify random effects for both of them; if not, only specify the random effects for the one that appear in the data.  Likewise, any IV that does not appear in the dataset should not be included anywhere in the formula.

For instance, had one of these datasets been the `lme4::sleepstudy` data, you would have replaced `ss_form <- NULL` with:

```{r ss-form, eval=FALSE}
ss_form <- Reaction ~ Days + (Days | Subject)
```

In other words: __all you need is the formula, NOT the call to `lme4::lmer()`; you are not actually fitting any model to the data.__

#### Dataset 1

Investigate the experimental design for `dataset1` and type the appropriate model formula.

`r hide()`

`DV ~ A * B * C + (A * B | subj_id)`

```{r ds1-sol}
ds1 %>% count(subj_id, A, B, C) %>% filter(subj_id == 1L)
```

`r unhide()`

#### Dataset 2

Investigate the experimental design for `dataset2` and type the appropriate model formula.

`r hide()`

```
DV ~ A * B * C + (A + B + C + A:B + A:C + B:C | subj_id) +
                 (A * B * C | item_id)
```

```{r ds2-sol}
ds2 %>% count(subj_id, A, B, C) %>% filter(subj_id == 1L)
```

#### Dataset 3

Investigate the experimental design for `dataset3` and type the appropriate model formula.

`r hide()`

`DV ~ A * B + (A | subj_id) + (B | item_id)`

```{r ds3-sol1}
ds3 %>% count(subj_id, A, B) %>% filter(subj_id == 1L)
```

```{r ds3-sol2}
ds3 %>% count(item_id, A, B) %>% filter(item_id == 1L)
```

`r unhide()`

#### Dataset 4

Investigate the experimental design for `dataset4` and type the appropriate model formula.

`r hide()`

`DV ~ A * B + (1 | subj_id) + (A * B | item_id)`

```{r ds4-sol1}
ds4 %>% count(subj_id, A, B) %>% filter(subj_id == 1L)
```

```{r ds4-sol2}
ds4 %>% count(item_id, A, B) %>% filter(item_id == 1L)
```

`r unhide()`

## Simulating data with crossed random factors

For this first set of exercises, we will generate simulated data corresponding to an experiment with a single, two-level factor (independent variable) that is within-subjects and between-items.  Let's imagine that the experiment involves lexical decisions to a set of words (e.g., is "PINT" a word or nonword?), and the dependent variable is response time (in milliseconds), and the independent variable is word type (noun vs verb).  We want to treat both subjects and words as random factors (so that we can generalize to the population of events where subjects encounter words).  You can play around with the web app below (or [click here to open it in a new window](http://shiny.psy.gla.ac.uk/Dale/crossed){target="_blank"}), which allows you to manipulate the data-generating parameters and see their effect on the data.

```{r webapp, echo=FALSE, fig.cap="*Web app for crossed random effects.*"}
knitr::include_app("http://shiny.psy.gla.ac.uk/Dale/crossed", height = "600px")
```

Here is the DGP for response time $Y_{si}$ for subject $s$ and item $i$:

*Level 1:*

\begin{equation}
Y_{si} = \beta_{0s} + \beta_{1} X_{i} + e_{si}
\end{equation}

*Level 2:*

\begin{equation}
\beta_{0s} = \gamma_{00} + S_{0s} + I_{0i}
\end{equation}

\begin{equation}
\beta_{1} = \gamma_{10} + S_{1s}
\end{equation}

*Variance Components:*

\begin{equation}
\langle S_{0s}, S_{1s} \rangle \sim N\left(0, \mathbf{\Sigma}\right) 
\end{equation}

\begin{equation}
\mathbf{\Sigma} = \left(\begin{array}{cc}{\tau_{00}}^2 & \rho\tau_{00}\tau_{11} \\
         \rho\tau_{00}\tau_{11} & {\tau_{11}}^2 \\
         \end{array}\right) 
\end{equation}

\begin{equation}
I_{0s} \sim N\left(0, {\omega_{00}}^2\right) 
\end{equation}

\begin{equation}
e_{si} \sim N\left(0, \sigma^2\right)
\end{equation}

In the above equation, $X_i$ is a numerical predictor coding which condition the item $i$ is in; e.g., -.5 for noun, .5 for verb.

We could just reduce levels 1 and 2 to 

$$Y_{si} = \beta_0 + S_{0s} + I_{0i} + (\beta_1 + S_{1s})X_{i} + e_{si}$$

where:

|Parameter    | Symbol| Description                                       |
|:------------|:------|:--------------------------------------------------|
| \(Y_{si}\)  | `Y`   | RT for subject \(s\) responding to item \(i\);    |
| \(\beta_0\) | `b0`  | grand mean;                                       |
| \(S_{0s}\)  | `S_0s` | random intercept for subject \(s\);               |
| \(I_{0i}\)  | `I_0i` | random intercept for item \(i\);                  |
| \(\beta_1\) | `b1` | fixed effect of word type (slope);                |
| \(S_{1s}\)  | `S_1s` | by-subject random slope;                          |
| \(X_{i}\)   | `cond` | deviation-coded predictor variable for word type; |
| \(\tau_{00}\) | `tau_00` | by-subject random intercept standard deviation |
| \(\tau_{11}\) | `tau_11` | by-subject random slope standard deviation |
| \(\rho\)    | `rho` | correlation between random intercept and slope |
| \(\omega_{00}\) | `omega_00` | by-item random intercept standard deviation |
| \(e_{si}\)  | `err` | residual error                                   |
| \(\sigma\)  | `sig` | residual error standard deviation                 |

### Set up the environment and define the parameters for the DGP

If you want to get the same results as everyone else for this exercise, then we all should seed the random number generator with the same value.  While we're at it, let's load in the packages we need.

```{r sim-setup, message=FALSE}
library("lme4")
library("tidyverse")

set.seed(11709)  
```

Now let's define the parameters for the DGP (data generating process).

```{r dgp-params}
nsubj <- 100 # number of subjects
nitem <- 50  # must be an even number

b0 <- 800 # grand mean
b1 <- 80 # 80 ms difference
effc <- c(-.5, .5) # deviation codes

omega_00 <- 80 # by-item random intercept sd (omega_00)

## for the by-subjects variance-covariance matrix
tau_00 <- 100 # by-subject random intercept sd
tau_11 <- 40 # by-subject random slope sd
rho <- .2 # correlation between intercept and slope

sig <- 200 # residual (standard deviation)
```

You'll create three tables:

| Name       | Description                                                          |
|:-----------|:---------------------------------------------------------------------|
| `subjects` | table of subject data including `subj_id` and subject random effects |
| `items`    | table of stimulus data including `item_id` and item random effect    |
| `trials`   | table of trials enumerating encounters between subjects/stimuli      |

Then you will merge together the information in the three tables, and calculate the response variable according to the model formula above.

** Generate a sample of stimuli

Let's randomly generate our `r nitem` items. Create a tibble called `item` like the one below, where `iri` are the by-item random intercepts (drawn from a normal distribution with variance \(\omega_{00}^2\) ` `iri_sd^2`).  Half of the words are of type NOUN (`cond` ` -.5) and half of type VERB (`cond` ` .5).

```{r item-tib}
items <- tibble(item_id = 1:nitem,
                cond = rep(c(-.5, .5), times = nitem / 2),
                I_0i = rnorm(nitem, 0, sd = omega_00))
```

```{r items}
items
```

`r hide("Hint for making cond")`

`rep()`

`r unhide()`

`r hide("Hint for making item random effects")`

`rnorm()`

`r unhide()`

`r hide("Solution")`

```{r item-tib-sol, eval=FALSE}
items <- tibble(item_id = 1:nitem,
                cond = rep(c(-.5, .5), times = nitem / 2),
                I_0i = rnorm(nitem, 0, sd = omega_00))

items
```

```{r items-print, echo=FALSE}
items
```

`r unhide()`

### Generate a sample of subjects

To generate the by-subject random effects, you will need to generate data from a *bivariate normal distribution*.  To do this, we will use the function `MASS::mvrnorm`.  

<div class="warning">

REMEMBER: do not run `library("MASS")` just to get this one function, because `MASS` has a function `select()` that will overwrite the tidyverse version. Since all we want from MASS is the `mvrnorm()` function, we can just access it directly by the `pkgname::function` syntax, i.e., `MASS::mvrnorm()`.

</div>

Your subjects table should look like this:

`r hide("Click to reveal full table")`

```{r subj-tbl1, echo = FALSE}
cov <- rho * tau_00 * tau_11

mx <- matrix(c(tau_00^2, cov,
               cov,      tau_11^2),
             nrow = 2)

by_subj_rfx <- MASS::mvrnorm(nsubj, mu = c(S_0s = 0, S_1s = 0), Sigma = mx)

subjects <- as_tibble(by_subj_rfx) %>%
  mutate(subj_id = row_number()) %>%
  select(subj_id, everything())

subjects %>% print(n = +Inf)
```

`r unhide()`

`r hide("Hint 1")`

recall that:

* *`tau_00`*: by-subject random intercept standard deviation
* *`tau_11`*: by-subject random slope standard deviation
* *`rho`* : correlation between intercept and slope

`r unhide()`

`r hide("Hint 2")`

`covariance = rho * tau_00 * tau_11`

`r unhide()`

`r hide("Hint 3")`

```{r hint3, eval=FALSE}
matrix(    tau_00^2,            rho * tau_00 * tau_11,
        rho * tau_00 * tau_11,      tau_11^2, ...)

```

`r unhide()`

`r hide("Hint 4")`

```{r hint4, eval=FALSE}
as_tibble(mx) %>%
  mutate(subj_id = ...)
```

`r unhide()`

`r hide("Solution")`

```{r code-subj-sol, eval=FALSE}
cov <- rho * tau_00 * tau_11

mx <- matrix(c(tau_00^2, cov,
               cov,      tau_11^2),
             nrow = 2)

by_subj_rfx <- MASS::mvrnorm(nsubj, mu = c(S_0s = 0, S_1s = 0), Sigma = mx)

subjects <- as_tibble(by_subj_rfx) %>%
  mutate(subj_id = row_number()) %>%
  select(subj_id, everything())

subjects %>% print(n = +Inf)
```

```{r print-subj, echo = FALSE}
subjects %>% print(n = +Inf)
```

`r unhide()`

### Generate a sample of encounters (trials)

Each trial is an *encounter* between a particular subject and stimulus.  In this experiment, each subject will see each stimulus.  Generate a table `trials` that lists the encounters in the experiments. Note: each participant encounters each stimulus item once.  Use the `crossing()` function to create all possible encounters.

Now apply this example to generate the table below, where `err` is the residual term, drawn from \(N \sim \left(0, \sigma^2\right)\), where \(\sigma\) is `err_sd`.

```{r gen-encounters, echo=FALSE}
trials <- crossing(subj_id = subjects %>% pull(subj_id),
                   item_id = items %>% pull(item_id)) %>%
  mutate(err = rnorm(nrow(.), mean = 0, sd = sig))

trials
```


`r hide()`

```{r trials2, eval=FALSE}
trials <- crossing(subj_id = subjects %>% pull(subj_id),
                   item_id = items %>% pull(item_id)) %>%
  mutate(err = rnorm(nrow(subjects) * nrow(items), mean = 0, sd = sig))
```

```{r trials2-print, echo=FALSE}
trials
```

`r unhide()`

### Join `subjects`, `items`, and `trials`

Merge the information in `subjects`, `items`, and `trials` to create the full dataset `dat_sim`, which looks like this:

```{r make-dat, echo=FALSE}
dat_sim <- subjects %>%
  inner_join(trials, "subj_id") %>%
  inner_join(items, "item_id") %>%
  arrange(subj_id, item_id) %>%
  select(subj_id, item_id, S_0s, I_0i, S_1s, cond, err)

dat_sim
```

`r hide()`

```{r dat-sim-sol, eval=FALSE}
dat_sim <- subjects %>%
  inner_join(trials, "subj_id") %>%
  inner_join(items, "item_id") %>%
  arrange(subj_id, item_id) %>%
  select(subj_id, item_id, S_0s, I_0i, S_1s, cond, err)

dat_sim
```

```{r sim-print, echo=FALSE}
dat_sim
```

`r unhide()`

### Create the response variable

Add the response variable `Y` to dat according to the model formula:

$$Y_{si} = \beta_0 + S_{0s} + I_{0i} + (\beta_1 + S_{1s})X_{i} + e_{si}$$

so that the resulting table (`dat_sim2`) looks like this:

```{r dat-sim2, echo=FALSE}
dat_sim2 <- dat_sim %>%
  mutate(Y = b0 + S_0s + I_0i + (S_1s + b1) * cond + err) %>%
  select(subj_id, item_id, Y, everything())

dat_sim2
```

Note: this is the full **decomposition table** for this model.

`r hide()`

```{r dat-sim2-sol, eval=FALSE}
dat_sim2 <- dat_sim %>%
  mutate(Y = b0 + S_0s + I_0i + (S_1s + b1) * cond + err) %>%
  select(subj_id, item_id, Y, everything())

dat_sim2
```

```{r dat-sim2-print, echo=FALSE}
dat_sim2
```

`r unhide()`

## Fitting the model

Now that you have created simulated data, estimate the model using `lme4::lmer()`, and run `summary()`.

`r hide()`

```{r fit-model}
mod_sim <- lmer(Y ~ cond + (1 + cond | subj_id) + (1 | item_id),
                dat_sim2, REML = FALSE)

summary(mod_sim, corr = FALSE)
```

`r unhide()`

Now see if you can identify the data generating parameters in the output of `summary()`.

```{r dgp1, include=FALSE}
srfx <- attr(VarCorr(mod_sim)$subj_id, "stddev")
irfx <- attr(VarCorr(mod_sim)$item_id, "stddev")
rc <- attr(VarCorr(mod_sim)$subj_id, "correlation")[1, 2]

res <- attr(VarCorr(mod_sim), "sc")

ffx <- fixef(mod_sim)
```

First, try to find \(\beta_0\) and \(\beta_1\).

`r hide("Solution: Fixed effects")`

```{r fef-sol, echo = FALSE}
tribble(~parameter, ~variable, ~input, ~estimate,
        "\\(\\hat{\\beta}_0\\)", "`b0`", b0, as.numeric(round(ffx[1], 3)),
        "\\(\\hat{\\beta}_1\\)", "`b1`", b1, as.numeric(round(ffx[2], 3))) %>%
  knitr::kable()
```

`r unhide()`

Now try to find estimates of random effects parameters \(\tau_{00}\), \(\tau_{11}\), \(\rho\), \(\omega_{00}\), and \(\sigma\).

`r hide("Solution: Random effects parameters")`

```{r rfx-sol, echo = FALSE}
tribble(~parameter, ~variable, ~input, ~estimate,
        "\\(\\hat{\\tau}_{00}\\)", "`tau_00`", tau_00,
        as.numeric(round(srfx[1], 3)),
        "\\(\\hat{\\tau}_{11}\\)", "`tau_11`", tau_11,
        as.numeric(round(srfx[2], 3)),
        "\\(\\hat{\\rho}\\)", "`rho`", rho, as.numeric(round(rc, 3)),
        "\\(\\hat{\\omega}_{00}\\)", "`omega_00`", omega_00,
        as.numeric(round(irfx[1], 3)),
        "\\(\\hat{\\sigma}\\)", "`sig`", sig,
        as.numeric(round(res, 3))) %>%
  knitr::kable()
```

`r unhide()`
